---
title: "Dynamic Visualization using Shiny"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)

# libraries
library( ggplot2 )

```

# Motivating question: Why do we want large sample sizes?

In statistics, you should have learned that when we sample, we want large sample sizes. *Why?*

```{r gplot-example-setup}

set.seed( 2026 )
sigma <- 1
n_reps <- 2000

df <- do.call( 
  rbind, 
  lapply( c(2, 10, 50), function( n ) {
    means <- replicate( n_reps, mean( rnorm( n, mean = 0, sd = sigma ) ) )
    data.frame( n = as.factor( n ), sample_mean = means )
    }
    ) 
  )

```

We can visualize this with a series of plots:

```{r gplot-example}

ggplot( 
  df, 
  aes( x = sample_mean, fill = n ) ) +
  geom_histogram( bins = 30, alpha = 0.6, position = "identity" ) +
  facet_wrap( ~ n, nrow = 1 ) +
  labs( 
    title = "Sampling distribution of the sample mean", 
    x = "Sample mean", y = "" ) +
  theme_minimal() +
  theme( legend.position = "none" )

```

Take a look at each plot. What is the visualization showing you?

Let's take a look at a different way of using a visualization to convey this point. Take a look at this [visualization of sampling ditributions](https://jacobtnyoung.shinyapps.io/sampling-distributions/){target="_blank" rel="noopener"}.


## Other examples...

[network data collection tool(https://jacobtnyoung.shinyapps.io/probation-pilot-app/){target="_blank" rel="noopener"}
Network disruption app


## Static vs. Dynamic Visualization

The visualizations above are examples of static and dynamic visualizations, respectively. What makes the latter visualization *dynamic* is that the user can alter the inputs to change what the visualization shows. It is not simply fixed by the author/developer/etc, it is interactive: the **user** is able to alter what is *rendered*.

# What is Shiny and How it Works

Shiny is an R framework for building interactive web applications directly from R code. It allows you to take static analyses and visualizations, like a `ggplot`, and turn them into dynamic, user-driven experiences that run in a web browser. 

With Shiny, users can:

  - Adjust inputs (sliders, dropdowns, buttons)
  - Filter data
  - Change model parameters
  - Instantly see updated visualizations and results

## Core Concepts

Shiny apps are built around three simple ideas:

  - The UI (User Interface), which is what the user sees. It includes the layout of the page, input controls (sliders, dropdowns, checkboxes), and output placeholders (where plots, tables, or text will appear). The UI defines the structure of the app.
  
  - The Server, this is what happens behind the scenes. It is where we put the instructions for what is produced in the UI, where data is processed, and where calculations happen. The server connects user inputs to outputs.
  
  - Reactivity is what makes Shiny dynamic (and powerful). Reactivity means that code reruns automatically when inputs change. Think of it like Excel cell recalculation: If one cell changes, any dependent cells update automatically. Shiny works the same way in that changes to the inputs trigger update which refresh the output.

Let's go back to the [visualization of sampling ditributions](https://jacobtnyoung.shinyapps.io/sampling-distributions/){target="_blank" rel="noopener"} app and take a closer look. 

## An Example

### Static Visualization

Suppose we created a plot using this code:

```r

library( ggplot2 )

n <- 100

x <- rnorm( n )

ggplot(data.frame(x = x), aes(x = x)) +
      geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
      labs(title = paste("Histogram of", n, "random normals"),
           x = "value", y = "count") +
      theme_minimal()

```

We have a very simple plot of random draws from a standard normal distribution. What if we wanted to increase the number we took? Decrease it?

### Dynamic Visualization

Let's create a very simple Shiny app. As we do this, ask yourself: What is part of the UI? What is part of the server? Where is reactivity happening?

To get started, create a new R script file and add this code to it:

```r

library( shiny )
library( ggplot2 )

ui <- fluidPage(
  titlePanel("Interactive Histogram"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("n", "Sample size (n):", min = 10, max = 1000, value = 100, step = 10)
    ),
    mainPanel(
      plotOutput("hist")
    )
  )
)

server <- function(input, output) {
  output$hist <- renderPlot({
    x <- rnorm(input$n)
    ggplot(data.frame(x = x), aes(x = x)) +
      geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
      labs(title = paste("Histogram of", input$n, "random normals"),
           x = "value", y = "count") +
      theme_minimal()
  })
}

shinyApp(ui = ui, server = server)

```

Ok, you just created your first Shiny app! I knew you had it in you. Let's walk through what we did.

The first part loads the libraries to get the functions needed to build the app and the plot:

```r

library( shiny )
library( ggplot2 )

```

Next, we have our two key pieces, the `ui` and the `server`. Let's look at the code for defining the user interface:

```r

ui <- fluidPage(
  titlePanel("Interactive Histogram"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("n", "Sample size (n):", min = 10, max = 1000, value = 100, step = 10)
    ),
    mainPanel(
      plotOutput("hist")
    )
  )
)

```

We are first defining an object called `ui` which will define the user interface. Look at your app and see what the user side features are. Can you find them in the code?  

We start the user interface with the `fluidPage()` function. Inside this function we have `titlePanel()` which gives the title. 

We then we have `sidebarLayout()` which prepares the sidebar. Inside the sidebar layout we have `sidebarPanel()` that has `sliderInput()` inside it. What does this line do? Look at the app and think about what is happening.

Finally, we have the `mainPanel()` function with the `plotOutput()` function designating an object `"hist"` to be plotted.


Now, let's examine the server side:

```r 

server <- function(input, output) {
  output$hist <- renderPlot({
    x <- rnorm(input$n)
    ggplot(data.frame(x = x), aes(x = x)) +
      geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
      labs(title = paste("Histogram of", input$n, "random normals"),
           x = "value", y = "count") +
      theme_minimal()
  })
}

```

We define a **function** called `server` which is going to be our workhorse. It is going to take an `input` that the user can alter in the UI. Anytime a change is made by the user, this is going to pass through our server object and change what is rendered in `output`.

Remember that we defined an object called `"hist"` in UI using `plotOutput()`. Here we are defining what that will be. Specifically, we use the `renderPlot()` function. This is getting a bit more complicated, so let's take each line:

  - `x <- rnorm(input$n)` creates an object `x`, which is a random draw from a normal distribution, `rnorm`, where the size is defined by the user with `input$n`. Look back to the UI, where do you see `"n"`? **Look at what we just did**: we built into our plot function the changes that are being made by the user in the UI.
  - `ggplot(data.frame(x = x), aes(x = x)) +`... defines a plot object, using `ggplot()`, that is based on the `x` object.

And that is really it. Much of the code is tweaking the `ggplot` object.

Now for our last piece:

```r

shinyApp(ui = ui, server = server)

```

Here, we are just running the `shinyApp()` function and defining the arguments that function needs to operate: the UI and the server.


# Using AI to Build Shiny Apps


# Exercise


# Q&A and Extensions

Here are some really good FREE books:

  - [Mastering Shiny](https://mastering-shiny.org/index.html) is the first stop for moving forward with learning how to use Shiny.
  
  https://unleash-shiny.rinterface.com/
  https://engineering-shiny.org/index.html













